RDF Validation Requirements - Evaluation and Logical Underpinning

### 2015-01-06 paper/RDF validation requirements/RDF validation requirements.tex
5.2 but need to modify it again

\begin{itemize}
	\item R-21-IRI-PATTERN-MATCHING-ON-RDF-SUBJECTS
  \item R-22-IRI-PATTERN-MATCHING-ON-RDF-OBJECTS
  \item R-23-IRI-PATTERN-MATCHING-ON-RDF-PROPERTIES
\end{itemize}

\begin{itemize}
	\item IRI pattern matching on subjects, properties, and objects
\end{itemize}

@@

\begin{itemize}
  \item R-44-PATTERN-MATCHING-ON-RDF-LITERALS
\end{itemize}

\begin{itemize}
	\item pattern matching on literals
\end{itemize}

@@

\begin{ex}
# OWL 2 DL (functional-style syntax)
# ----------------------------------
Declaration( Datatype( :SSN ) ) 
DatatypeDefinition( 
    :SSN
    DatatypeRestriction( xsd:string xsd:pattern "[0-9]{3}-[0-9]{2}-[0-9]{4}" ) )     
DataPropertyRange( :hasSSN :SSN ) 
\end{ex}

\begin{ex}
# OWL 2 DL (turtle syntax)
# ------------------------
:SSN 
    a rdfs:Datatype ;
    owl:equivalentClass [
        a rdfs:Datatype ;
        owl:onDatatype xsd:string ;
        owl:withRestrictions ( 
            [ xsd:pattern "[0-9]{3}-[0-9]{2}-[0-9]{4}" ] ) ] .
:hasSSN rdfs:range :SSN .
\end{ex}

@@

Luke's droids can only have the numbers "R2-D2" or "C-3PO".
The universal restriction part of this constraint can be expressed by OWL 2 DL:
\ms{LukesDroids $\sqsubseteq$ $\forall$ droidNumber . DroidNumber}.
The restriction of the datatype \ms{DroidNumber}, however, cannot be expressed in DL, but OWL 2 DL can be used anyway:

@@

\begin{itemize}
  \item R-44-PATTERN-MATCHING-ON-RDF-LITERALS
\end{itemize}

\begin{itemize}
	\item negative pattern matching on literals
\end{itemize}

\begin{ex}
# examples
# --------
1. dbo:isbn format is different ’!’ from '^([iIsSbBnN 0-9-])*$'
2. dbo:postCode format is different ‘!’ from 'ˆ[0-9]{5}$'
3. foaf:phone contains any letters ('[A-Za-z]')
\end{ex}

@@

\begin{ex}
# test binding (DQTP)
# -------------------
dbo:isbn format is different ’!’ from '^([iIsSbBnN 0-9-])*$'

@@

\begin{ex}
# DQTP
# ----
SELECT DISTINCT ?s WHERE { ?s %%P1%% ?value .
    FILTER ( %%NOP%% regex(str(?value), %%REGEX%) ) }
\end{ex}

@@

\begin{ex}
# valid data
# ----------
:FoundationsOfSWTechnologies
    dbo:isbn 'ISBN-13 978-1420090505' .
\end{ex}

\begin{ex}
# invalid data
# ------------
:HandbookOfSWTechnologies
    dbo:isbn 'DOI 10.1007/978-3-540-92913-0' .
\end{ex}

@@

\begin{itemize}
	\item R-43-LITERAL-VALUE-COMPARISON
\end{itemize}

\begin{itemize}
	\item dbo:deathDate before ‘<’ dbo:birthDate
  \item dbo:releaseDate after ‘>’ dbo:latestReleaseDate
  \item dbo:demolitionDate before ‘<’ dbo:buildingStartDate
\end{itemize}

@@

\begin{ex}
# DSCL (DQTP)
# -----------
SELECT ?s WHERE { 
    ?s %%P1%% ?v1 .
    ?s %%P2%% ?v2 .
    FILTER ( ?v1 %%OP%% ?v2 ) }
\end{ex}

COMP Pattern \cite{Kontokostas2014}

Depending on the property semantics,
there are cases where two different literal values must have
a specific ordering with respect to an operator. 
P1 and P2 are the datatype properties we need to compare and 
OP is the comparison operator (\textless, \textless=, \textgreater, \textgreater=, =, !=).

\begin{ex}
# test binding (DQTP)
# -------------------
dbo:deathDate before ‘<’ dbo:birthDate

@@

\begin{ex}
# valid data
# ----------
:AlbertEinstein
    dbo:birthDate '1879-03-14'^^xsd:date ;
    dbo:deathDate '1955-04-18'^^xsd:date .
\end{ex}

\begin{ex}
# invalid data
# ------------
:NeilArmstrong
    dbo:birthDate '2012-08-25'^^xsd:date ;
    dbo:deathDate '1930-08-05'^^xsd:date .
\end{ex}

@@

duplicate strings are not allowed:

@@

\begin{itemize}
	\item R-142-NEGATIVE-RANGES-OF-RDF-LITERAL-VALUES
\end{itemize}

\begin{itemize}
	\item The literal value of a resource (having a certain type) must (not) be within a specific range
  \item P1 is a data property of an instance of class T1 and its literal value must be between the range of [Vmin,Vmax] or outside ( ‘!' ).
\end{itemize}

@@

\begin{itemize}
	\item R-45-RANGES-OF-RDF-LITERAL-VALUES
\end{itemize}

@@

\begin{ex}
# OWL 2 DL (functional-style syntax)
# ----------------------------------
Declaration( Datatype( :NumberPlayersPerWorldCupTeam ) ) 
DatatypeDefinition( 
    :NumberPlayersPerWorldCupTeam
    DatatypeRestriction( 
        xsd:nonNegativeInteger 
        xsd:minInclusive "1"^^xsd:nonNegativeInteger 
        xsd:maxInclusive "23"^^xsd:nonNegativeInteger ) )     
DataPropertyRange( :position :NumberPlayersPerWorldCupTeam ) 
\end{ex}

\begin{ex}
# OWL 2 DL (turtle syntax)
# ------------------------
:NumberPlayersPerWorldCupTeam
    a rdfs:Datatype ;
    owl:equivalentClass [
        a rdfs:Datatype ;
        owl:onDatatype xsd:nonNegativeInteger ;
        owl:withRestrictions ( 
            [ xsd:minInclusive "1"^^xsd:nonNegativeInteger ]
            [ xsd:maxInclusive "23"^^xsd:nonNegativeInteger ] ) ] .
:position rdfs:range :NumberPlayersPerWorldCupTeam .
\end{ex}

\begin{itemize}
	\item The data range 'NumberPlayersPerWorldCupTeam' contains the non negative integers 1 to 23, as each world cup team can only have 23 football players at most
\end{itemize}

\begin{ex}
# valid data
# ----------
:MarioGoetze
    :position "19"^^:NumberPlayersPerWorldCupTeam .
\end{ex}

\begin{ex}
# invalid data
# ------------
:MarioGoetze
    :position "99"^^:NumberPlayersPerWorldCupTeam .
\end{ex}

\begin{itemize}
	\item range of position is the datatype 1-23
  \item constraints are stated on this datatype
\end{itemize}

@@

\begin{itemize}
  \item R-121-SPECIFY-ORDER-OF-RDF-RESOURCES
	\item R-217-DEFINE-ORDER-FOR-FORMS/DISPLAY
\end{itemize}

\begin{itemize}
	\item order properties
	\item order objects of object properties
	\item order literal of data properties
\end{itemize}

>>>>

\ms{IRI Pattern Matching} corresponds to the requirements
\begin{itemize}
	\item \ms{R-21-IRI-PATTERN-MATCHING-ON-RDF-SUBJECTS},
	\item \ms{R-22-IRI-PATTERN-MATCHING-ON-RDF-OBJECTS} and
	\item \ms{R-23-IRI-PATTERN-MATCHING-ON-RDF-PROPERTIES} 
\end{itemize}
indicating IRI pattern matching on subjects, properties, and objects.

@@

\ms{Literal Pattern Matching} corresponds to the requirement \\
\ms{R-44-PATTERN-MATCHING-ON-RDF-LITERALS}.
indicating pattern matching on literals.

@@

\begin{ex}
# OWL 2 DL (functional-style syntax):
Declaration( Datatype( SSN ) ) 
DatatypeDefinition( 
    SSN
    DatatypeRestriction( xsd:string xsd:pattern "[0-9]{3}-[0-9]{2}-[0-9]{4}" ) )     
DataPropertyRange( hasSSN SSN ) 
\end{ex}

\begin{ex}
# OWL 2 DL (turtle syntax):
SSN 
    a rdfs:Datatype ;
    owl:equivalentClass [
        a rdfs:Datatype ;
        owl:onDatatype xsd:string ;
        owl:withRestrictions ( 
            [ xsd:pattern "[0-9]{3}-[0-9]{2}-[0-9]{4}" ] ) ] .
hasSSN rdfs:range SSN .
\end{ex}

A social security number is a string that matches the given regular expression. 
The second axiom defines SSN as an abbreviation for a datatype restriction on xsd:string. 
The first axiom explicitly declares SSN to be a datatype. 
The datatype SSN can be used just like any other datatype; 
for example, it is used in the third axiom to define the range of the hasSSN property.

\begin{ex}
# valid data:
TimBernersLee
    hasSSN "123-45-6789"^^SSN .
\end{ex}

\begin{ex}
# invalid data:
TimBernersLee
    hasSSN "123456789"^^SSN .
\end{ex}

@@

Luke's droids can only have the numbers "R2-D2" or "C-3PO".
The universal restriction part of this constraint can be expressed by OWL 2 DL:
\ms{LukesDroids $\sqsubseteq$ $\forall$ droidNumber.DroidNumber}.
The restriction of the datatype \ms{DroidNumber}, however, cannot be expressed in DL, but OWL 2 DL can be used anyway:

@@

\ms{Negative Literal Pattern Matching} corresponds to the requirement \\
\ms{R-44-PATTERN-MATCHING-ON-RDF-LITERALS}
indicating negative pattern matching on literals.

\begin{ex}
# examples:
1. dbo:isbn format is different ’!’ from '^([iIsSbBnN 0-9-])*$'
2. dbo:postCode format is different ‘!’ from 'ˆ[0-9]{5}$'
3. foaf:phone contains any letters ('[A-Za-z]')
\end{ex}

@@

\begin{ex}
# test binding (DQTP):
dbo:isbn format is different ’!’ from '^([iIsSbBnN 0-9-])*$'

@@

\begin{ex}
# DQTP:
SELECT DISTINCT ?s WHERE { ?s %%P1%% ?value .
    FILTER ( %%NOP%% regex(str(?value), %%REGEX%) ) }
\end{ex}

@@

\begin{ex}
# valid data:
FoundationsOfSWTechnologies
    dbo:isbn 'ISBN-13 978-1420090505' .
\end{ex}

\begin{ex}
# invalid data:
HandbookOfSWTechnologies
    dbo:isbn 'DOI 10.1007/978-3-540-92913-0' .
\end{ex}

@@

\ms{Literal Value Comparison} corresponds to the requirement \\
\ms{R-43-LITERAL-VALUE-COMPARISON}.
Examples are:

\begin{itemize}
	\item dbo:deathDate before ‘\textless’ dbo:birthDate
  \item dbo:releaseDate after ‘\textgreater’ dbo:latestReleaseDate
  \item dbo:demolitionDate before ‘\textless’ dbo:buildingStartDate
\end{itemize}

@@

\begin{ex}
# DSCL (DQTP):
SELECT ?s WHERE { 
    ?s %%P1%% ?v1 .
    ?s %%P2%% ?v2 .
    FILTER ( ?v1 %%OP%% ?v2 ) }
\end{ex}

This constraint corresponds to the COMP Pattern \cite{Kontokostas2014}. 
Depending on the property semantics,
there are cases where two different literal values must have
a specific ordering with respect to an operator. 
P1 and P2 are the datatype properties we need to compare and 
OP is the comparison operator (\textless, \textless=, \textgreater, \textgreater=, =, !=).

\begin{ex}
# test binding (DQTP):
dbo:deathDate before ‘<’ dbo:birthDate

@@

\begin{ex}
# valid data:
:AlbertEinstein
    dbo:birthDate '1879-03-14'^^xsd:date ;
    dbo:deathDate '1955-04-18'^^xsd:date .
\end{ex}

\begin{ex}
# invalid data:
:NeilArmstrong
    dbo:birthDate '2012-08-25'^^xsd:date ;
    dbo:deathDate '1930-08-05'^^xsd:date .
\end{ex}

@@

Duplicate strings are not allowed:

@@

\ms{Negative Literal Ranges} corresponds to the requirement
\ms{R-142-NEGATIVE-} \ms{RANGES-OF-RDF-LITERAL-VALUES}.
The literal value of a resource (having a certain type) must (not) be within a specific range.
P1 is a data property of an instance of class T1 and its literal value must be between the range of [Vmin,Vmax] or outside ( ‘!' ).

@@

\ms{Literal Ranges} corresponds to the requirement
\ms{R-45-RANGES-OF-RDF-LITERAL-} \ms{VALUES}.

@@

\begin{ex}
# OWL 2 DL (functional-style syntax):
Declaration( Datatype( NumberPlayersPerWorldCupTeam ) ) 
DatatypeDefinition( 
    NumberPlayersPerWorldCupTeam
    DatatypeRestriction( 
        xsd:nonNegativeInteger 
        xsd:minInclusive "1"^^xsd:nonNegativeInteger 
        xsd:maxInclusive "23"^^xsd:nonNegativeInteger ) )     
DataPropertyRange( position NumberPlayersPerWorldCupTeam ) 
\end{ex}

\begin{ex}
# OWL 2 DL (turtle syntax):
NumberPlayersPerWorldCupTeam
    a rdfs:Datatype ;
    owl:equivalentClass [
        a rdfs:Datatype ;
        owl:onDatatype xsd:nonNegativeInteger ;
        owl:withRestrictions ( 
            [ xsd:minInclusive "1"^^xsd:nonNegativeInteger ]
            [ xsd:maxInclusive "23"^^xsd:nonNegativeInteger ] ) ] .
position rdfs:range NumberPlayersPerWorldCupTeam .
\end{ex}

The data range 'NumberPlayersPerWorldCupTeam' contains the non negative integers 1 to 23, as each world cup team can only have 23 football players at most.

\begin{ex}
# valid data:
MarioGoetze
    position "19"^^:NumberPlayersPerWorldCupTeam .
\end{ex}

\begin{ex}
# invalid data:
MarioGoetze
    position "99"^^:NumberPlayersPerWorldCupTeam .
\end{ex}

%The range of \ms{position} is the datatype 1-23
  %\item constraints are stated on this datatype
%\end{itemize}

@@

\ms{Define Order} corresponds to the requirements
\ms{R-121-SPECIFY-ORDER-OF-RDF-} \ms{RESOURCES} and
\ms{R-217-DEFINE-ORDER-FOR-FORMS/DISPLAY}.
With this constraint objects of object properties can be ordered as well as literals of data properties and objects of object properties.
